---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---

```{r}
# load data set
heartattack <- read.csv("heart_attack_prediction_dataset.csv", header=T)

# Our population of interest are people at risk of heart attack 
heartattack <- heartattack[heartattack$Heart.Attack.Risk == 1,]
head(heartattack)
```

### Find recommended sample size for this study

```{r}
# calculate min sample size needed
pop_size <- nrow(heartattack) # 3139

# using 95% CI, find n for worst case scenario: p = 0.5
MOE <- 0.05
z <- 1.96
p_guess <- 0.5

# if N is large enough to ignore FPC
n_0 = ceiling( ((2*z)^2*(0.5)*(0.5)) / (MOE^2)) # 1537
# since we know N = 8763, using FPC
n = ceiling( n_0 / (1 + (n_0/pop_size)) ) # 1032
```

Assuming the worst case proportions 0.5, the sample size used if we ignored FPC is `r n_0`. Whereas including FPC the sample size used in SRS will be `r n`.

### Compare study design for stratification

#### Method 1: stratify by sex

```{r}
#Calculate within variance of each sex: Male, Female
variance_within_strata <- aggregate(BMI ~ Sex, heartattack, var)
colnames(variance_within_strata) <- c("Sex","Within Variance Sex")
print(variance_within_strata)

#Get stratum sizes
male_stratum_size <- nrow(heartattack[heartattack$Sex == "Male",])
female_stratum_size <- nrow(heartattack[heartattack$Sex == "Female",])


#Sample size n_h proportional to N_h*S_pw^2/sqrt(cost)
#Ignore costs
total <- sum(male_stratum_size*variance_within_strata$`Within Variance Sex`[1],
            female_stratum_size*variance_within_strata$`Within Variance Sex`[2])

male_size_proportion <- 
  male_stratum_size*variance_within_strata$`Within Variance Sex`[1]/total

female_size_proportion <- 
  female_stratum_size*variance_within_strata$`Within Variance Sex`[2]/total

male_sample_size <- round(male_size_proportion*n)
female_sample_size <- round(female_size_proportion*n)

#Overall stratified variance
var.strata <- c(variance_within_strata$`Within Variance Sex`[1], 
                variance_within_strata$`Within Variance Sex`[2])
wt.strata <- c(male_size_proportion, female_size_proportion)

overall.sex.var <- sum(wt.strata*var.strata)
print(overall.sex.var)
```

#### Method 2: stratify by diet

```{r}
#Calculate within variance of each diet stratum: Average, Unhealthy, Healthy
variance_within_strata <- aggregate(BMI ~ Diet, heartattack, var)
colnames(variance_within_strata) <- c("Diet","Within Variance BMI")
variance_within_strata

#Get stratum sizes
average_stratum_size <- nrow(heartattack[heartattack$Diet == "Average",])
healthy_stratum_size <- nrow(heartattack[heartattack$Diet == "Healthy",])
unhealthy_stratum_size <- nrow(heartattack[heartattack$Diet == "Unhealthy",])


#Sample size n_h proportional to N_h*S_pw^2/sqrt(cost)
#Ignore costs
total <- sum(average_stratum_size*variance_within_strata$`Within Variance BMI`[1],
            healthy_stratum_size*variance_within_strata$`Within Variance BMI`[2],
            unhealthy_stratum_size*variance_within_strata$`Within Variance BMI`[3])

average_size_proportion <- 
  average_stratum_size*variance_within_strata$`Within Variance BMI`[1]/total
healthy_size_proportion <- 
  healthy_stratum_size*variance_within_strata$`Within Variance BMI`[2]/total
unhealthy_size_proportion <- 
  unhealthy_stratum_size*variance_within_strata$`Within Variance BMI`[3]/total

average_sample_size <- round(average_size_proportion*n)
healthy_sample_size <- round(healthy_size_proportion*n)
unhealthy_sample_size <- round(unhealthy_size_proportion*n)

#Overall stratified variance
var.strata <- c(variance_within_strata$`Within Variance BMI`[1], 
                variance_within_strata$`Within Variance BMI`[2],
                variance_within_strata$`Within Variance BMI`[3])
wt.strata <- 
  c(average_size_proportion, healthy_size_proportion, unhealthy_size_proportion)

overall.diet.var <- sum(wt.strata*var.strata)
print(overall.diet.var)
```

#### Method 3: stratify by whether patient has diabetes

```{r}
#Calculate within variance of whether patient has diabetes: 1: Yes, 0: No
variance_within_strata <- aggregate(BMI ~ Diabetes, heartattack, var)
colnames(variance_within_strata) <- c("Diabetes","Within Variance Diabetes")
print(variance_within_strata)

#Get stratum sizes
diabetes_stratum_size <- nrow(heartattack[heartattack$Diabetes == 1,])
no_diabetes_stratum_size <- nrow(heartattack[heartattack$Diabetes == 0,])


#Sample size n_h proportional to N_h*S_pw^2/sqrt(cost)
#Ignore costs
total <- 
  sum(diabetes_stratum_size*variance_within_strata$`Within Variance Diabetes`[1],
      no_diabetes_stratum_size*variance_within_strata$`Within Variance Diabetes`[2])

diabetes_size_proportion <-
  diabetes_stratum_size*variance_within_strata$`Within Variance Diabetes`[1]/total
no_diabetes_size_proportion <- 
  no_diabetes_stratum_size*variance_within_strata$`Within Variance Diabetes`[2]/total

diabetes_sample_size <- round(diabetes_size_proportion*n)
no_diabetes_sample_size <- round(no_diabetes_size_proportion*n)

#Overall stratified variance
var.strata <- c(variance_within_strata$`Within Variance Diabetes`[1], 
                variance_within_strata$`Within Variance Diabetes`[2])
wt.strata <- c(diabetes_size_proportion, no_diabetes_size_proportion)

overall.diabetes.var <- sum(wt.strata*var.strata)
print(overall.diabetes.var)
```

#### Method 4: stratify by whether patient has family history of heart-related problems

```{r}
#Calculate within variance of whether patient has 
#family history of heart-related problems:#1: Yes, 0: No

variance_within_strata <- aggregate(BMI ~ Family.History, heartattack, var)
colnames(variance_within_strata) <- c("Family History","Within Variance Family History")
print(variance_within_strata)

#Get stratum sizes
history_stratum_size <- nrow(heartattack[heartattack$Family.History == 1,])
no_history_stratum_size <- nrow(heartattack[heartattack$Family.History == 0,])


#Sample size n_h proportional to N_h*S_pw^2/sqrt(cost)
#Ignore costs
total <- 
  sum(history_stratum_size*variance_within_strata$`Within Variance Family History`[1],
   no_history_stratum_size*variance_within_strata$`Within Variance Family History`[2])

history_size_proportion <- 
  history_stratum_size*variance_within_strata$`Within Variance Family History`[1]/total
no_history_size_proportion <- 
  no_history_stratum_size*variance_within_strata$`Within Variance Diabetes`[2]/total

history_sample_size <- round(history_size_proportion*n)
no_history_sample_size <- round(no_history_size_proportion*n)

#Overall stratified variance
var.strata <- c(variance_within_strata$`Within Variance Family History`[1], 
                variance_within_strata$`Within Variance Family History`[2])
wt.strata <- c(history_size_proportion, no_history_size_proportion)

overall.history.var <- sum(wt.strata*var.strata)
print(overall.history.var)
```

#### Method 5: stratify by obesity status

```{r}
#Calculate within variance of obesity status: 1: Obese, 0: Not obese
variance_within_strata <- aggregate(BMI ~ Obesity, heartattack, var)
colnames(variance_within_strata) <- c("Obesity","Within Variance Obesity")
print(variance_within_strata)

#Get stratum sizes
obesity_stratum_size <- nrow(heartattack[heartattack$Obesity == 1,])
not_obese_stratum_size <- nrow(heartattack[heartattack$Obesity == 0,])


#Sample size n_h proportional to N_h*S_pw^2/sqrt(cost)
#Ignore costs
total <- sum(obesity_stratum_size*variance_within_strata$`Within Variance Obesity`[1],
            not_obese_stratum_size*variance_within_strata$`Within Variance Obesity`[2])

obesity_size_proportion <- 
  obesity_stratum_size*variance_within_strata$`Within Variance Obesity`[1]/total
not_obese_size_proportion <- 
  not_obese_stratum_size*variance_within_strata$`Within Variance Obesity`[2]/total

history_sample_size <- round(obesity_size_proportion*n)
no_history_sample_size <- round(not_obese_size_proportion*n)

#Overall stratified variance
var.strata <- c(variance_within_strata$`Within Variance Obesity`[1], 
                variance_within_strata$`Within Variance Obesity`[2])
wt.strata <- c(obesity_size_proportion, not_obese_size_proportion)

overall.obesity.var <- sum(wt.strata*var.strata)
print(overall.obesity.var)
```

```{r}
overall_var <- 
  data.frame(overall.sex.var, 
             overall.diet.var,
             overall.diabetes.var, 
             overall.history.var, 
             overall.obesity.var)

colnames(overall_var) <-
  c("Overall Sex Var.", 
    "Overall Diet Var.", 
    "Overall Diabetes Var.", 
    "Overall History Var.", 
    "Overall Obesity Var.")

print(overall_var)
```

By computing and comparing the within variances based on different stratas, stratifying by sex gave the lowest overall within variance of 39.09994. Since the stratification study design performs the best for the largest between-strata variance, implying the lowest within-strata variance, we will stratify by sex.

In the two stratums: Sex = (Male, Female), sample size for Male is `r male_sample_size` and sample size for Female is `r female_sample_size`

```         
Selecting Samples through SRS and Stratification by sex
```

```{r}
# set seed
set.seed(2023)

# take SRS of n = 1032
SRS.index <- sample.int(pop_size, n, replace=F)
SRS_sample <- heartattack[SRS.index, ]
head(SRS_sample)

#Stratify male and female stratums to take samples from
male_stratum <- heartattack[heartattack$Sex == "Male",]
female_stratum <- heartattack[heartattack$Sex == "Female",]

#Take Stratified samples of males (n = 708) and females (n = 324)
stratified_male.index <- sample.int(male_stratum_size, male_sample_size, replace = F)
male_sample <- male_stratum[stratified_male.index,]
head(male_sample)
nrow(male_sample)

stratified_female.index <- sample.int(female_stratum_size, female_sample_size, replace = F)
female_sample <- female_stratum[stratified_female.index,]
head(female_sample)
nrow(female_sample)
```

Calculating Estimates

```{r}
#Calculate mean BMI from SRS

SRS_BMI_mean <- mean(SRS_sample$BMI)

#Calculate mean BMI from male sample and female sample

male_BMI_mean <- mean(male_sample$BMI)
female_BMI_mean <- mean(female_sample$BMI)
#Calculate stratified estimator for BMI mean (sum of weighted BMI means)

strata_estimator_BMI_mean <- (male_stratum_size/pop_size)*male_BMI_mean +
                                 (female_stratum_size/pop_size)*female_BMI_mean


data.frame(`Sampling Method` = c("SRS","Stratified Estimate"),
           `BMI Mean` = c(SRS_BMI_mean,strata_estimator_BMI_mean))
```

Calculate standard error

```{r}
#Calculate SE for SRS and Stratified

#SRS SE calculation
SRS_variance <- sum((SRS_sample$BMI - SRS_BMI_mean)^2)/(n-1)
SRS_FPC <- (1- n/pop_size)
SRS_SE <- sqrt(SRS_FPC * SRS_variance/n)

#Stratified SE calculation
male_strata_variance <- sum((male_sample$BMI - male_BMI_mean)^2)/(male_sample_size-1)
male_strata_FPC <- (1 - male_sample_size/male_stratum_size)
male_proportion_squared <- (male_stratum_size/pop_size)^2

female_strata_variance <- 
  sum((female_sample$BMI - female_BMI_mean)^2)/(female_sample_size-1)
female_strata_FPC <- (1 - female_sample_size/female_stratum_size)
female_proportion_squared <- (female_stratum_size/pop_size)^2

stratified_SE <- sqrt(
  (male_proportion_squared*male_strata_FPC*male_strata_variance/male_sample_size)+
(female_proportion_squared*female_strata_FPC*female_strata_variance/female_sample_size))

data.frame(`Sampling Method` = c("SRS","Stratification"),
           SE = c(SRS_SE,stratified_SE))
```
